\section{Introduction}
\label{sec:introduction}

Water pollution threatens ecosystems and public health worldwide, yet existing
monitoring systems rely on sparse, single-modality measurements that detect
contamination only after significant damage has occurred. The fundamental challenge
is that no single sensing modality can detect all pollution types early enough:
chemical sensors measure only pre-specified analytes, satellite remote sensing
cannot resolve sub-surface processes, and biological indicators integrate over
unknown timescales.

Biology, by contrast, runs a massively redundant detection system. Molecular
stress responses activate within minutes of toxicant exposure. Organismal
behavior changes before physiological damage occurs. Microbial communities shift
in composition as environmental conditions change. Spectral properties of water
bodies reflect integrated biogeochemical state. Each channel captures different
aspects of water quality with different latencies and specificities. The question
is: can we build an AI system that reads all these channels simultaneously?

Existing approaches in AI-driven water quality monitoring operate within
individual modalities. HydroGEM~\citep{Loreaux2025} applies foundation models to
sensor time series; Zhi et al.~\citep{Zhi2024} demonstrate satellite-based
prediction of non-optically-active parameters; and scattered work applies
machine learning to environmental metagenomics. However, no system integrates
multiple modalities, and no architecture addresses the core technical
challenges of environmental multimodal fusion: asynchronous sampling,
heterogeneous mathematical spaces, routinely missing modalities, and cascade
decision-making.

In this work, we present \textbf{SENTINEL} (\textbf{S}ynergistic
\textbf{E}nvironmental \textbf{N}etwork for \textbf{T}emporal
\textbf{I}ntelligence and \textbf{N}etworked \textbf{E}cological
\textbf{L}earning), the first multimodal AI framework for early water
pollution detection. Our contributions are:

\begin{itemize}
    \item \textbf{SENTINEL-DB}: A large-scale multimodal water quality dataset
    harmonizing 390M+ records from thirteen public sources across five modality
    types, with a unified parameter ontology and H3 spatial cross-modal linking.

    \item \textbf{Five novel architectures}: AquaSSM (continuous-time SSM for
    sensor data), HydroViT (water-specific satellite foundation model),
    MicroBiomeNet (Aitchison-geometry-aware metagenomics encoder), ToxiGene
    (pathway-informed transcriptomic classifier), and BioMotion
    (diffusion-pretrained behavioral anomaly detector).

    \item \textbf{SENTINEL-Fusion}: A Perceiver IO-derived cross-modal temporal
    attention framework with temporal decay attention and confidence-weighted
    gating, designed for asynchronous heterogeneous environmental monitoring.

    \item \textbf{Theoretical contributions}: Cross-modal transfer bounds,
    compositional neural network guarantees, and distribution-free conformal
    anomaly detection with coverage guarantees.

    \item \textbf{A public platform}: A web application enabling anyone to
    assess water quality via GPS, photos, or test kit readings, with
    automated citizen data quality control.
\end{itemize}
