\section{Methods}
\label{sec:methods}

SENTINEL processes five environmental data modalities through
modality-specific encoders that each project to a shared 256-dimensional
embedding space, then fuses them via a Perceiver IO cross-modal temporal
attention framework. We describe each component below.

\subsection{AquaSSM: Continuous-Time State Space Model}
\label{sec:aquassm}

Water quality sensor networks produce multivariate time series at irregular
intervals. AquaSSM extends the selective state space mechanism~\citep{Gu2024}
to continuous time.

\paragraph{Continuous-Time Selective State Space.}
Instead of discretizing the state transition matrix $\mathbf{A}$ with a fixed
step size, AquaSSM parameterizes the step size as a learned function of the
actual time gap:
\begin{equation}
    \Delta t_k = f_\theta(t_k - t_{k-1}, \mathbf{x}_{k-1})
\end{equation}
where $f_\theta$ is a small MLP with softplus output. The state update becomes:
\begin{equation}
    \mathbf{h}_k = e^{\mathbf{A} \cdot \Delta t_k} \mathbf{h}_{k-1}
    + \mathbf{B} \cdot \Delta t_k \cdot \mathbf{x}_k
\end{equation}
with diagonal $\mathbf{A}$ stored as $\log(-\mathbf{A})$ for numerical stability.

\paragraph{Multi-Scale Temporal Decomposition.}
Eight parallel SSM channels with characteristic timescales from 1 hour to 365
days capture patterns at different temporal resolutions. A gated mixing layer
combines their outputs:
\begin{equation}
    \mathbf{y} = \sum_{s=1}^{8} g_s \cdot \mathbf{y}_s, \quad
    g_s = \sigma(\mathbf{W}_g [\mathbf{y}_1; \ldots; \mathbf{y}_8])_s
\end{equation}

\paragraph{Physics-Informed Constraints.}
Soft constraints encode known relationships as auxiliary losses:
DO solubility decreasing with temperature (Benson-Krause formula),
pH-alkalinity coupling through carbonate chemistry, and
conductivity-TDS proportionality. These act as inductive biases
during training.

\paragraph{Sensor Health Sentinel.}
A dedicated head classifies each sensor into five health states
(normal, drift, fouling, failure, calibration needed) from
reconstruction error statistics, enabling automatic down-weighting
of unreliable sensor contributions.

\subsection{HydroViT: Satellite Foundation Model}
\label{sec:hydrovit}

HydroViT pretrains a ViT-Base backbone exclusively on water pixels
from Sentinel-2 and Sentinel-3, using masked autoencoder (MAE)
reconstruction~\citep{He2022} augmented with a spectral physics
consistency loss ensuring water-leaving radiance constraints.

\paragraph{Multi-Resolution Fusion.}
Sentinel-2 (10m, 5-day) and Sentinel-3 OLCI (300m, daily) are fused
through bidirectional cross-attention. S3 provides temporal context;
S2 provides spatial detail.

\paragraph{16-Parameter Output.}
A multi-head regression network jointly predicts 16 water quality
parameters with heteroscedastic aleatoric uncertainty estimates.

\subsection{MicroBiomeNet: Compositional Metagenomics Encoder}
\label{sec:microbiomenet}

Microbiome data lies on the Aitchison simplex, not in Euclidean
space~\citep{GordonRodriguez2022}. MicroBiomeNet operates natively
in this geometry.

\paragraph{Aitchison-Aware Attention.}
All internal representations use centered log-ratio (CLR) coordinates.
Attention similarity is computed in Aitchison geometry:
\begin{equation}
    \text{sim}(\mathbf{q}, \mathbf{k}) =
    \frac{\text{clr}(\mathbf{q})^\top \text{clr}(\mathbf{k})}
    {\|\text{clr}(\mathbf{q})\| \|\text{clr}(\mathbf{k})\|}
\end{equation}

\paragraph{Simplex Neural ODE.}
For longitudinal samples, community dynamics are modeled as a neural
ODE in the Aitchison tangent space, enabling trajectory-based anomaly
detection.

\subsection{ToxiGene: Pathway-Informed Transcriptomic Classifier}
\label{sec:toxigene}

ToxiGene mirrors the biological hierarchy
(gene $\to$ pathway $\to$ process $\to$ outcome) with sparse,
biology-constrained connections from Reactome and AOP-Wiki.
An information bottleneck discovers minimal diagnostic gene panels.

\subsection{BioMotion: Behavioral Anomaly Detection}
\label{sec:biomotion}

BioMotion pretrains trajectory encoders by learning to denoise
corrupted movement trajectories~\citep{Guo2024}, using the
denoising score as an anomaly signal. Multi-organism ensemble
(Daphnia, mussel, fish) captures different contaminant sensitivities.

\subsection{SENTINEL-Fusion}
\label{sec:fusion}

\paragraph{Perceiver IO with Temporal Memory.}
A fixed array of $N=256$ learned latent vectors serves as a
compressed ``state of the waterway.'' New modality embeddings
update this array through cross-attention.

\paragraph{Temporal Decay Attention.}
Attention weights are modulated by learned per-modality-pair
exponential decay rates, capturing that behavioral data stales
in minutes while microbial data remains relevant for weeks.

\paragraph{Confidence-Weighted Gating.}
Each encoder outputs a calibrated confidence score. The fusion
layer gates contributions accordingly, automatically handling
unreliable inputs without explicit rules.

\paragraph{Cascading Alert Heads.}
Four output heads: anomaly detection, contaminant classification,
source attribution, and escalation recommendation.
