\section{Related Work}
\label{sec:related_work}

\paragraph{AI for Water Quality Monitoring.}
Deep learning has been applied to individual water quality modalities.
Zhi et al.~\citep{Zhi2024} demonstrated that satellite imagery can predict
non-optically-active parameters through learned correlations.
Loreaux et al.~\citep{Loreaux2025} introduced HydroGEM, a foundation model
for sensor time series using TCN-Transformer architectures.
Ghattas et al.~\citep{Ghattas2024} applied foundation models to harmful
algal bloom detection from satellite data. These approaches are
single-modality; SENTINEL is the first to unify all five modalities.

\paragraph{State Space Models.}
Structured state space models~\citep{Gu2024} achieve linear-time sequence
modeling. Mamba extends this with selective scan mechanisms.
Ansari et al.~\citep{Ansari2025} adapt foundation models (Chronos-2) for
time series forecasting. AquaSSM extends these to continuous time with
physics-informed constraints specific to water quality.

\paragraph{Multimodal Fusion.}
Perceiver IO~\citep{Jaegle2021} handles heterogeneous inputs through
iterative cross-attention. CLIP~\citep{Radford2021} learns shared
representations across modalities. However, no existing architecture
handles the specific combination of asynchronous sampling, heterogeneous
mathematical spaces, and routinely missing modalities present in
environmental monitoring.

\paragraph{Compositional Data Analysis.}
Aitchison geometry provides the correct mathematical framework for
compositional data~\citep{GordonRodriguez2022}. We extend this to
deep learning architectures with provable compositional coherence.

\paragraph{Earth Observation Foundation Models.}
Prithvi-EO-2.0~\citep{Jakubik2024} and Clay provide general-purpose
Earth observation models. HydroViT differs by pretraining exclusively
on water pixels with water-specific spectral constraints.

\paragraph{Behavioral Ecotoxicology.}
Commercial biomonitor systems (DaphTox, MFB) use statistical thresholds
for behavioral anomaly detection. Guo et al.~\citep{Guo2024} introduced
diffusion pre-training for fish trajectory recognition. BioMotion
extends this to multi-organism ensemble with learned anomaly fusion.
